|

### 2023-01-16 inst/MyArticle/paper.Rmd
cleanup figures

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column making it easy to integrate new inputs in *prolfqua* if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in *MSstats*[@Choi2014] format, only a few lines of code are needed to update the *prolfqua* `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences among the various input formats in column names and enables the use of the methods without additional arguments. An example code for creating an *FragPipe*[@yu2020fast] configuration can be found in the Supplementary Section "Creating a prolfqua configuration". We implemented methods that transform the data into tidy data for popular software like *MaxQuant*[@Cox2008MaxQuant], or *FragPipe*, which stores the same variable, e.g., intensity, in multiple columns, one for each sample. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base *R* [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features (see Supplementary Figure 8) in *RStudio* [@RStudio2022], and help to implement argument free functions.

@@

```{r LFQData, echo=FALSE, fig.cap="(ref:LFQData)", out.width = '90%'}
knitr::include_graphics("LFQData_UML_LARGE.png")
```

@@

knitr::include_graphics("ContrastClassesUML_LARGE.png")
```

@@

The package *prolfqua* provides functions to determine the vector of $parameter$ weights $c$, from a linear model and a contrast specification string. In the Supplement we provided an example, how to specify contrasts for a dataset with two explanatory variables and an interaction term (Supplementary Section "Specifying Contrasts for Models with two Factors and Interaction Term").

@@

However, if there is no observation in a group the model fit fails. For example, suppose a protein is unobserved in all the samples of a group. In that case, a plausible explanation is that the protein abundance is below the limit of detection (LOD) of the MS instrument. In such a case, we will substitute the group mean using the expected protein abundance $A$ at the LOD $A_{LOD}$.
To estimate $A_{LOD}$ we are using the protein abundances of those groups where the protein was observed in only a single sample (see Supplementary Section "Estimating $A_{LOD}$"). Typically there are many such cases, and hence we take the median.

@@

Using the data frame of tidy data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")` [@quast2022protti]. To simplify the integration of *prolfqua* with Bioconductor based workflows there is a method that converts the `LFQData` class into a `r Biocpkg("SummarizedExperiment")`. The use of R6 classes, which encapsulate the configuration and the data, allows for writing very concise code where functions can have few arguments. Auto-completion support for R6 classes in the *RStudio* editor makes it easy for novices to explore *prolfqua*'s functionality (see Supplementary Figure 8).

@@

However, computing the t-statistics at the peptide level and then summarizing it for each protein using their median produces the highest $pAUC_{10}$ scores among all the tested models (see Figure \@ref(fig:figure3) Panel B *prolfqua_ropeca*). Furthermore, by using the Beta distribution to model the number of peptides observed, we can further improve the $pAUC$ scores (see Figure \@ref(fig:figure3) Panel B center). However, the properties of Beta-based probabilities need to be better understood; their distribution is not uniform under the null hypothesis (see Supplementary Section "The probabilities produced by ROPECA are not p-values"). Therefore, the resulting $FDR$ estimates are biased (see Figure \@ref(fig:figure3) Panel C). Consequently, we can not recommend this method if an unbiased estimate of $FDR$ is essential, which is frequently the case. In addition, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset using this method (see Figure \@ref(fig:figure3) Panel C).

@@

We obtained difference and FDR estimates for all proteins and comparisons, as shown in Figure \@ref(fig:figure3) Panel A when using: a) the probabilistic dropout model (*proDA*), b) the hurdle model (*msqrobHurdle*), and c) *prolfqua_merged*. We observe that the performance of the scaled $p$-values or the $t$-statistics is comparable among these three methods (Figure \@ref(fig:figure3) Panel B). We tested if there is a significant difference between the $pAUC_{10}$ for all three methods, but did not reject the null hypothesis that there is no such difference (Supplementary Section "DEA benchmark : IonStar/MaxQuant/peptide.txt - Significance test"). Also, the FDR estimates (Figure \@ref(fig:figure3) Panel C) are comparable for all three methods.

Furthermore, all three models perform similarly when examined using a different benchmark dataset CPTAC/MaxQuant (see Methods). For this dataset, *proDA* performed slightly but not significantly better than *prolfqua* and *msqrobHurdle* (and Supplementary Section "DEA benchmark : CPTAC/MaxQuant/peptide.txt").

In addition, we examined the DEA performance when using protein intensities reported by quantification software *FragPipe* for the *IonStar* dataset as input. Using protein abundances as input significantly simplifies the analysis and interpretation and might benefit from optimization implemented in the quantification software. However, we can only fit the *proDA* and *prolfqua_merged* (see Table \@ref(tab:describeBenchmarkedModels)) models to protein abundances, while *MSstats* and *msqrobHurdle* require psm or peptide level abundances In this DEA benchmark, *prolfqua* performed slightly, but not significantly better than *proDA* (Supplementary Section "DEA benchmark : IonStar/FragPipeV14/combined_protein.tsv").

Finally, we also compared the DEA performances when starting the analysis from the precursor abundances reported in the 'MSstats.tsv' file, generated by *FragPipe* v14, from the *IonStar* dataset. Since *MSstats*, *msqrob2*, *proDA*, and *prolfqua* all read `MStstats.tsv` files, we could eliminate a confounding factor, i.e., different input abundances (Supplementary Section "DEA benchmark: IonStar/FragPipeV14/MSstats.tsv"). In this DEA benchmark, *msqrob2* and *prolfqua_merged* perform best but not significantly better than *proDA* or *MSstats*. Furthermore, by processing the *IonStar* dataset with both *MaxQuant* and *FragPipe*, we can compare their performances (Supplementary Section "Comparing DEA results for *MaxQuant* and *FragPipe*").

We focused our benchmark on comparing the statistical modeling methods while we fixed the pre-processing steps. However, some of these steps are of utmost significance when performing differential expression analysis [@frohlich2022benchmarking]. One of them is the normalization of the abundances within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer proteins from peptide identifications [@nesvizhskii2005interpretation] and protein abundances from peptide abundances is an additional important factor [@Grossmann2010]. For instance, the original *proDA* publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ abundances reported by *MaxQuant*, the $pAUC_{10}$ is lower ($pAUC_{10}(\textrm{t-statistics}) = 66\%$) compared with results obtained when protein abundances are estimated from peptide abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 72\%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast] used to identify and quantify proteins significantly contributes to the entire pipeline's performance altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis. In the Supplementary Section "Comparing DEA results for *MaxQuant* and *FragPipe*" we compare DEA benchmarking results for the quantification software. While the number of proteins identified with two peptides is practically the same, the DEA benchmark performance differs significantly by $\approx 10\%$ for the $pAUC_{10}$ score. This difference is more significant than the differences due to the choice of the modeling method.

@@

@@

file.path("hexStickerProlfqua.png") |>
  knitr::include_graphics()
```

>>>>

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column making it easy to integrate new inputs in *prolfqua* if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in *MSstats*[@Choi2014] format, only a few lines of code are needed to update the *prolfqua* `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences among the various input formats in column names and enables the use of the methods without additional arguments. An example code for creating an *FragPipe*[@yu2020fast] configuration can be found in the Supporting Material S3 Section "Creating a prolfqua configuration". We implemented methods that transform the data into tidy data for popular software like *MaxQuant*[@Cox2008MaxQuant], or *FragPipe*, which stores the same variable, e.g., intensity, in multiple columns, one for each sample. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base *R* [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features (see Supporting Figure 8) in *RStudio* [@RStudio2022], and help to implement argument free functions.

@@

```{r LFQData, echo=FALSE, fig.cap="(ref:LFQData)", out.width = '90%'}
knitr::include_graphics("graphics/LFQData_UML_LARGE.png")
```

@@

knitr::include_graphics("graphics/ContrastClassesUML_LARGE.png")
```

@@

The package *prolfqua* provides functions to determine the vector of $parameter$ weights $c$, from a linear model and a contrast specification string. In the Supporting Material S10 we provided an example, how to specify contrasts for a dataset with two explanatory variables and an interaction term.

@@

However, if there is no observation in a group the model fit fails. For example, suppose a protein is unobserved in all the samples of a group. In that case, a plausible explanation is that the protein abundance is below the limit of detection (LOD) of the MS instrument. In such a case, we will substitute the group mean using the expected protein abundance $A$ at the LOD $A_{LOD}$.
To estimate $A_{LOD}$ we are using the protein abundances of those groups where the protein was observed in only a single sample (see Supporting Material S8 "Estimating $A_{LOD}$"). Typically there are many such cases, and hence we take the median.

@@

Using the data frame of tidy data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")` [@quast2022protti]. To simplify the integration of *prolfqua* with Bioconductor based workflows there is a method that converts the `LFQData` class into a `r Biocpkg("SummarizedExperiment")`. The use of R6 classes, which encapsulate the configuration and the data, allows for writing very concise code where functions can have few arguments. Auto-completion support for R6 classes in the *RStudio* editor makes it easy for novices to explore *prolfqua*'s functionality (see Supporting Figure 8).

@@

However, computing the t-statistics at the peptide level and then summarizing it for each protein using their median produces the highest $pAUC_{10}$ scores among all the tested models (see Figure \@ref(fig:figure3) Panel B *prolfqua_ropeca*). Furthermore, by using the Beta distribution to model the number of peptides observed, we can further improve the $pAUC$ scores (see Figure \@ref(fig:figure3) Panel B center). However, the properties of Beta-based probabilities need to be better understood; their distribution is not uniform under the null hypothesis (see Supporting Material S9 "The probabilities produced by ROPECA are not p-values"). Therefore, the resulting $FDR$ estimates are biased (see Figure \@ref(fig:figure3) Panel C). Consequently, we can not recommend this method if an unbiased estimate of $FDR$ is essential, which is frequently the case. In addition, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset using this method (see Figure \@ref(fig:figure3) Panel C).

@@

We obtained difference and FDR estimates for all proteins and comparisons, as shown in Figure \@ref(fig:figure3) Panel A when using: a) the probabilistic dropout model (*proDA*), b) the hurdle model (*msqrobHurdle*), and c) *prolfqua_merged*. We observe that the performance of the scaled $p$-values or the $t$-statistics is comparable among these three methods (Figure \@ref(fig:figure3) Panel B). We tested if there is a significant difference between the $pAUC_{10}$ for all three methods, but did not reject the null hypothesis that there is no such difference (Supporting Material S3 "DEA benchmark : IonStar/MaxQuant/peptide.txt - Significance test"). Also, the FDR estimates (Figure \@ref(fig:figure3) Panel C) are comparable for all three methods.

Furthermore, all three models perform similarly when examined using a different benchmark dataset CPTAC/MaxQuant (see Methods). For this dataset, *proDA* performed slightly but not significantly better than *prolfqua* and *msqrobHurdle* (and Supporting Material S4 "DEA benchmark : CPTAC/MaxQuant/peptide.txt").

In addition, we examined the DEA performance when using protein intensities reported by quantification software *FragPipe* for the *IonStar* dataset as input. Using protein abundances as input significantly simplifies the analysis and interpretation and might benefit from optimization implemented in the quantification software. However, we can only fit the *proDA* and *prolfqua_merged* (see Table \@ref(tab:describeBenchmarkedModels)) models to protein abundances, while *MSstats* and *msqrobHurdle* require psm or peptide level abundances In this DEA benchmark, *prolfqua* performed slightly, but not significantly better than *proDA* (Supporting Material S5 "DEA benchmark : IonStar/FragPipeV14/combined_protein.tsv").

Finally, we also compared the DEA performances when starting the analysis from the precursor abundances reported in the 'MSstats.tsv' file, generated by *FragPipe* v14, from the *IonStar* dataset. Since *MSstats*, *msqrob2*, *proDA*, and *prolfqua* all read `MStstats.tsv` files, we could eliminate a confounding factor, i.e., different input abundances (Supporting Material S6 "DEA benchmark: IonStar/FragPipeV14/MSstats.tsv"). In this DEA benchmark, *msqrob2* and *prolfqua_merged* perform best but not significantly better than *proDA* or *MSstats*. Furthermore, by processing the *IonStar* dataset with both *MaxQuant* and *FragPipe*, we can compare their performances (Supporting Material S7 "Comparing DEA results for *MaxQuant* and *FragPipe*").

We focused our benchmark on comparing the statistical modeling methods while we fixed the pre-processing steps. However, some of these steps are of utmost significance when performing differential expression analysis [@frohlich2022benchmarking]. One of them is the normalization of the abundances within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer proteins from peptide identifications [@nesvizhskii2005interpretation] and protein abundances from peptide abundances is an additional important factor [@Grossmann2010]. For instance, the original *proDA* publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ abundances reported by *MaxQuant*, the $pAUC_{10}$ is lower ($pAUC_{10}(\textrm{t-statistics}) = 66\%$) compared with results obtained when protein abundances are estimated from peptide abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 72\%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast] used to identify and quantify proteins significantly contributes to the entire pipeline's performance altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis. In the Supporting Material S7 "Comparing DEA results for *MaxQuant* and *FragPipe*" we compare DEA benchmarking results for the quantification software. While the number of proteins identified with two peptides is practically the same, the DEA benchmark performance differs significantly by $\approx 10\%$ for the $pAUC_{10}$ score. This difference is more significant than the differences due to the choice of the modeling method.

@@

# Supporting Information

- Supporting Material S1: How to install *prolfqua* and *prolfquabenchmark*
- Supporting Material S2: Benchmark Vignettes
- Supporting Material S3: DEA benchmark IonStar/MaxQuant/peptide.txt - Significance test
- Supporting Material S4: DEA benchmark : CPTAC/MaxQuant/peptide.txt
- Supporting Material S5: DEA benchmark : IonStar/FragPipeV14/combined_protein.tsv
- Supporting Material S6: DEA benchmark : IonStar/FragPipeV14/MSstats.tsv
- Supporting Material S7: Comparing DEA results for *MaxQuant* and *FragPipe*
- Supporting Material S8: Estimating $A_{LOD}$
- Supporting Material S9: The probabilities produced by ROPECA are not p-values
- Supporting Material S10: Specifying Contrasts for Models with two Factors and Interaction Term
- Supporting Material S11: Creating a prolfqua configuration

@@

file.path("graphics/hexStickerProlfqua.png") |>
  knitr::include_graphics()
```

### 2023-01-16 inst/MyArticle/prolfqua_supplement.Rmd
cleanup figures

---
title: | 
    | Supporting Information
    | prolfqua : A Comprehensive R-package for Proteomics Differential Expression Analysis
author:
  - Witold E. Wolski:
      email: wew@fgcz.ethz.ethz.ch
      institute: [fgcz, sib]
      correspondence: true
  - Paolo Nanni:
      institute: [fgcz]
  - Jonas Grossmann:
      institute: [fgcz, sib]
  - Maria dErrico:
      institute: [fgcz, sib]
  - Ralph Schlapbach:
      institute: [fgcz]
  - Christian Panse:
      institute: [fgcz, sib]
institute:
  - fgcz: Functional Genomics Center Zurich (FGCZ), ETH Zurich / University of Zurich, Winterthurerstrasse 190, 8057 Zurich, Switzerland
  - sib: Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland
package: prolfqua
output:
  bookdown::pdf_document2:
    toc: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
abstract: |
  This document contains the supplements for the prolfqua manuscript available
  through \url{https://doi.org/10.1101/2022.06.07.494524}.
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
  bibliography: prolfqua.bib 
---

@@

# How to install *prolfqua* and *prolfquabenchmark*

@@

# DEA benchmark : IonStar/MaxQuant/peptide.txt - Significance test

@@

## Benchmark Vignettes

Pre-build version of the vignettes, we created to run the DEA benchmark, are available:

# DEA benchmark : CPTAC/MaxQuant/peptide.txt

@@

# DEA benchmark : IonStar/FragPipeV14/combined_protein.tsv

@@

# DEA benchmark : IonStar/FragPipeV14/MSstats.tsv

@@

# Comparing DEA results for *MaxQuant* and *FragPipe*

@@

# Estimating $A_{LOD}$

@@

# The probabilities produced by ROPECA are not p-values

@@

# Specifying Contrasts for Models with two Factors and Interaction Term

@@

# Creating a prolfqua configuration

@@

```{r proLFQuaSticker, echo=TRUE, out.height="5cm", eval=TRUE, fig.cap="Sticker maintainer: Witold E. Wolski; License: Creative Commons Attribution CC-BY. Feel free to share and adapt, but don't forget to credit the author."}
file.path("graphics/hexStickerProlfqua.png") |>
  knitr::include_graphics()
```

>>>>

---
title: | 
    | Supporting Information
    | prolfqua : A Comprehensive R-package for Proteomics Differential Expression Analysis
author:
  - Witold E. Wolski:
      email: wew@fgcz.ethz.ethz.ch
      institute: [fgcz, sib]
      correspondence: true
  - Paolo Nanni:
      institute: [fgcz]
  - Jonas Grossmann:
      institute: [fgcz, sib]
  - Maria dErrico:
      institute: [fgcz, sib]
  - Ralph Schlapbach:
      institute: [fgcz]
  - Christian Panse:
      institute: [fgcz, sib]
institute:
  - fgcz: Functional Genomics Center Zurich (FGCZ), ETH Zurich / University of Zurich, Winterthurerstrasse 190, 8057 Zurich, Switzerland
  - sib: Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland
package: prolfqua
output:
  bookdown::pdf_document2:
    toc: yes
    number_sections: false
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
abstract: |
  This document contains the supplements for the prolfqua manuscript available
  through \url{https://doi.org/10.1101/2022.06.07.494524}.
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
  bibliography: prolfqua.bib 
---

@@

# Material S1: How to install *prolfqua* and *prolfquabenchmark*

@@

# Material S2: Benchmark Vignettes (IonStar/MaxQuant)

Pre-build version of the vignettes, created to run the DEA benchmark, are available:

# Material S3: DEA benchmark IonStar/MaxQuant/peptide.txt - Significance test

@@

# Material S4: DEA benchmark : CPTAC/MaxQuant/peptide.txt

@@

# Material S5: DEA benchmark : IonStar/FragPipeV14/combined_protein.tsv

@@

# Material S6: DEA benchmark : IonStar/FragPipeV14/MSstats.tsv

@@

# Material S7: Comparing DEA results for *MaxQuant* and *FragPipe*

@@

# Material S8: Estimating $A_{LOD}$

@@

# Material S9: The probabilities produced by ROPECA are not p-values

@@

# Material S10: Specifying Contrasts for Models with two Factors and Interaction Term

@@

# Material S11: Creating a prolfqua configuration

@@

```{r proLFQuaSticker, echo=TRUE, out.height="5cm", eval=TRUE, fig.cap="Sticker maintainer: Witold E. Wolski; License: Creative Commons Attribution CC-BY. Feel free to share and adapt, but don't forget to credit the author."}
file.path("graphics/hexstrickerPRL2.png") |>
  knitr::include_graphics()
```
